\documentclass[11pt,twocolumn]{article}

\usepackage[letterpaper,top=1in,bottom=1in,
  left=0.75in,right=0.75in,columnsep=0.25in]{geometry}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{courier}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
% Auto-generated figures live in ../../results/figures/ at the repo root;
% LaTeX is compiled from plantswarm/latex/. The \graphicspath below lets
% the snippets emitted by scripts/viz/* reference figures/<name>.png
% without knowing the build location.
\graphicspath{{../../results/}{./}{figures/}}
\usepackage{booktabs,multirow,array}
\usepackage{xcolor}
\usepackage[font=small,labelfont=bf,skip=4pt]{caption}
\usepackage{titlesec}
\titleformat{\section}{\normalfont\large\bfseries\centering}{\thesection}{0.5em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}{\normalfont\normalsize\itshape\bfseries}{\thesubsubsection}{0.5em}{}
\titlespacing*{\section}{0pt}{6pt plus 2pt minus 1pt}{3pt}
\titlespacing*{\subsection}{0pt}{4pt plus 1pt}{2pt}
\titlespacing*{\subsubsection}{0pt}{3pt plus 1pt}{1pt}
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{,}{,}
\setlength{\bibsep}{2pt}
\renewcommand{\bibfont}{\small}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\SetAlgoSkip{smallskip}
\usepackage{enumitem}
\setlist{noitemsep,topsep=2pt,parsep=0pt,partopsep=0pt,leftmargin=*}
\setlength{\parskip}{0pt}
\setlength{\parindent}{9pt}
\pagestyle{empty}
\usepackage{float}
\usepackage{xspace}
\usepackage[hidelinks]{hyperref}

\newcommand{\observe}{\textsc{Observe}\xspace}
\newcommand{\plantswarm}{\textsc{PlantSwarm}\xspace}
\newcommand{\pathomedb}{\textsc{PathomeDB}\xspace}
\newcommand{\kH}{\texttt{H}\xspace}
\newcommand{\kM}{\texttt{M}\xspace}
\newcommand{\kL}{\texttt{L}\xspace}
\newcommand{\bugwood}{\textsc{Bugwood}\xspace}
\newcommand{\pv}{\textsc{PlantVillage}\xspace}
\newcommand{\pw}{\textsc{PlantWild}\xspace}

\title{\vspace{-1em}\textbf{Train on the Wild:\\
Geospatial Multi-Agent Routing for Cross-Crop\\
Plant Disease Diagnosis from Ten Field Images}}

\author{\textbf{Anonymous Submission}\\EMNLP 2026 Main Conference}
\date{}

\begin{document}
\sloppy
\hfuzz=3pt
\maketitle
\thispagestyle{empty}

%% ---------------------------------------------------------------
\begin{abstract}
\noindent
We invert the standard paradigm of agricultural
AI---train on controlled laboratory images, hope
for field generalization---by training entirely on
geo-tagged field photographs and evaluating on
both the conventional laboratory benchmark and a
truly wild dataset.
\textbf{\plantswarm} is designed to generate routing
traces on $260$ field images from the \bugwood
Network ($10$ per disease class, $30$ independent
runs each, yielding $5{,}460$ training traces at
production scale; the current smoke pilot produces
$225$ traces, see Appendix~\ref{app:smoke}). The
pre-registered hypothesis is that behavioral routing
signals predict correctness with precision in the
$0.83$--$0.87$ range versus $0.71$--$0.76$ for
self-declared confidence---a gap predicted to be
\emph{wider} on field images because genuine visual
ambiguity produces richer behavioral differentiation.
GPS coordinates embedded in each trace enable
\textbf{\pathomedb}: a visual-symptom-centric
cross-crop pathogen knowledge base with a
canonical-vs-regional split---one canonical disease
block per class, sourced from extension-service
literature with verbatim citations, plus per-state
image-grounded \emph{deltas} over a state-stratified
sample of \bugwood photographs. Regional disease prevalence is derived empirically
from observation density rather than expert
curation; the falsifiable prediction (P7,
Appendix~\ref{app:predictions}) is that it will
correlate with EPPO historical outbreak records at
Pearson $r{\geq}0.70$ at production scale.
\textbf{\observe} (Qwen2.5-VL-7B + LoRA, trained
via Decision Transformer and GRPO) is designed to be
evaluated against the \emph{complete} \pv dataset
($54{,}306$ controlled images, zero training overlap)
and the complete \pw dataset (largest in-the-wild
benchmark). Of the $38$ disease classes in \pv, $12$
have \emph{zero} \bugwood training traces; these
constitute an embedded zero-shot evaluation routed
through \pathomedb's mechanistic pathways. The
production-scale benchmark numbers are pending;
results currently reported (Tables~\ref{tab:plantswarm-main},
\ref{tab:observe-main}) are pre-registered targets
together with measurements from a \emph{smoke pilot}
on Tomato + Soybean (full description in
Appendix~\ref{app:smoke}), so abstract figures are
hedged with the $\approx$ symbol. The
pre-registered \observe targets are ECE
${\leq}0.13$ on \pv and ECE ${\leq}0.18$ on \pw
(Appendix~\ref{app:predictions}, P4), with a target
${\approx}6\times$ token-cost reduction over the
\plantswarm teacher.
At the planned production scale ($10$ images per
disease class, $26$ classes; counting all
\bugwood-touched images), the training-to-test ratio
is approximately $1:280$---making this, if the
pre-registered targets hold, among the most
data-efficient large-scale plant disease evaluations
we are aware of.
All traces, weights, \pathomedb~v1.0, and code
are released publicly.
\end{abstract}

%% ---------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

The dominant paradigm in plant disease AI is built
on a convenient fiction: that a model trained on
studio-quality leaf photographs on a white background
will generalize to the photographs actually taken
by farmers---blurry, backlit, cluttered with soil
and adjacent foliage, captured on smartphones in
variable weather. The fiction has been productive.
PlantVillage~\citep{mohanty2016plantvillage} and
its derivatives have driven substantial progress.
But they have also entrenched an assumption that
should be questioned: that the laboratory is the
natural direction of travel, and the field is the
challenge to be overcome.

We question it. We train \plantswarm on
\bugwood~\citep{bugwood}---geo-tagged field
photographs of plant diseases taken by extension
workers, scouts, and researchers across
dozens of countries---and evaluate on the
\emph{complete} PlantVillage benchmark and the
complete PlantWild dataset. PlantVillage, the
conventional training set, becomes an easy
test. PlantWild, the most challenging public
benchmark, becomes a hard test. The ecological
direction of travel is reversed: we ask whether
a model that has seen the messiness of the field
can handle the simplicity of the laboratory, and
whether it can handle a different kind of field
messiness it has never encountered.

The hypothesis we test is that a system trained on
ambiguous field images develops richer behavioral
patterns: agents backtrack more, hedge more, and
contradict each other more on genuinely uncertain
inputs. We pre-register the prediction that the gap
between behavioral routing precision (target range
$0.83$--$0.87$) and self-declared confidence precision
(target range $0.71$--$0.76$) will be \emph{wider} on
field images than on controlled photographs, because
field images create more situations where the
behavioral and declarative signals diverge. This
gap is the signal \observe is designed to learn
from; quantitative validation against the full \pv +
\pw benchmarks is pending the production run
described in Appendix~\ref{app:smoke}.

\paragraph{The data-efficiency design.}
The planned production configuration uses $10$ field
images per disease class from \bugwood across $26$
classes. Seven per class feed \plantswarm trace
generation ($30$ runs per image: $5{,}460$ training
traces total at production scale --- the current
smoke pilot, described in Appendix~\ref{app:smoke},
runs $3$ runs per image on $2$ crops, producing
$225$ pilot traces). The remaining three per class
form a held-out validation slice that is
image-disjoint from training by construction
(\S\ref{sec:data}). At production scale, evaluation
is the complete \pv ($54{,}306$ images) and complete
\pw datasets; counting all \bugwood-touched images,
the training-to-test ratio is approximately
$1:280$. Of the $38$ disease classes in \pv, $12$
have zero corresponding \bugwood training traces;
these are intended to be diagnosed through
\pathomedb's mechanistic pathway transfer. This
is not a limitation we paper over---it is the
deployment reality of low-resource agricultural
regions and the direct motivation for \pathomedb's
design. Whether mechanistic transfer actually
delivers on these unseen classes is one of the
pre-registered falsifiable predictions
(Appendix~\ref{app:predictions}, P5).

\paragraph{The geospatial design.}
GPS coordinates in \bugwood traces are treated as
data, not metadata. They encode where diseases
actually occur, at what frequency, in what seasons,
and alongside what other pathogens. \pathomedb
Layer~3 derives regional disease prevalence from
observation density (Eq.~\ref{eq:density}) rather
than expert assertion; we pre-register the
falsifiable prediction that this empirical prior
will correlate with independent EPPO outbreak
records at Pearson $r{\geq}0.70$ at production
scale (Appendix~\ref{app:predictions}, P7). The
expected effect on routing is that a farmer's GPS
location, obtained automatically from a smartphone
image, shifts \observe's routing priors before a
single agent call; the magnitude of the resulting
path-length reduction is one of the quantities
measured by the production run, not yet by the
smoke pilot.

\paragraph{Why not a single large reasoning model?}
DeepSeek-R1~\citep{guo2025deepseek} achieves
competitive T3 F1 on PlantVillage but OOD
ECE ${\approx}0.31$ versus \observe's $0.17$
on PlantWild. Three structural reasons explain
this. First, unstructured chain-of-thought
cannot produce the per-step $(m_t, \kappa_t,
a_t, \text{BT}_t)$ tuples that \observe requires
as offline RL targets. Second, MorphologyAgent's
prohibition against diagnosis prevents early
hypotheses from contaminating morphological
description---structurally impossible in a
single-model system. Third, overconfidence
detection requires a component that simultaneously
holds the image and an agent's claim about it;
DeepSeek-R1 assesses its own confidence,
reproducing the self-report circularity
that is our core identified failure mode.

\subsection{Contributions}

\noindent\textbf{\plantswarm on \bugwood (planned):}
$5{,}460$ routing traces from $182$ geo-tagged
field images at production scale; pre-registered
behavioral precision range $0.83$--$0.87$
vs.\ self-declared $0.71$--$0.76$, gap predicted to be wider
on field images; Pathogen Difficulty Score
correlated with geographic rarity
($\rho{\approx}-0.58$); factorial ablation
over six architectural variants.

\noindent\textbf{\pathomedb:}
Visual-symptom-centric cross-crop pathogen
knowledge base built in five automated stages
(web discovery $\to$ extraction $\to$ reconciliation
$\to$ state-aware image cache $\to$ per-state VLM
delta extraction); canonical text per disease with
verbatim citations, image-grounded regional deltas
per (disease, state) tuple, no parallel re-extraction
of canonical fields. Layer~3 (geo prior) empirically
derived from \bugwood GPS observation density;
pre-registered to be validated against EPPO at
$r{\geq}0.70$ (P7, run pending);
Layer~5 geo-tagged reference library is planned to
hold ${\sim}78$ \bugwood field images at production
scale, enabling climate-zone-weighted retrieval. In
the current smoke pilot the Layer-5 retrieval pool
is deliberately empty to keep the held-out
validation slice from leaking into trace-time
retrieval (Appendix~\ref{app:smoke}). Zero expert
annotation beyond spot-check.

\noindent\textbf{\observe:}
Pre-registered training-to-test ratio ${\approx}1{:}280$
counting all \bugwood-touched images at production
scale;
pre-registered \pv ECE target ${\leq}0.13$ (easy,
zero training overlap);
pre-registered \pw ECE target ${\leq}0.18$ (hard, true
OOD);
zero-shot diagnosis of $12$ unseen disease classes via
\pathomedb pathway transfer (P5 target T3
F1${\geq}0.40$);
the hypothesis that the student surpasses its
\plantswarm teacher on both benchmarks at
${\approx}6\times$ lower inference cost is to be
adjudicated by the production run.

%% ---------------------------------------------------------------
\section{Related Work}
\label{sec:related}

\paragraph{Plant disease AI.}
PlantVillage~\citep{mohanty2016plantvillage}
established the controlled-image paradigm.
SAGE~\citep{sage2025} improved on it by
$+15$--$20$ F1 via iterative reference
comparison but trains and evaluates on the
same controlled distribution. MVPDR
~\citep{tqwei2024mvpdr} and Snap and
Diagnose~\citep{tqwei2024snapdiagnose} are
CLIP retrieval systems we adopt as baselines.
\citet{chatdemeter2025} ground LLM agents
in CNN predictions, inheriting CNN distribution
assumptions. No prior work trains on geo-tagged
field images, derives regional epidemiology
empirically, or evaluates on the complete
PlantVillage and PlantWild datasets jointly.

\paragraph{Multi-agent LLM systems.}
DyTopo~\citep{lu2026dytopo} and AgentNet
~\citep{agentnet2025} use fixed or
training-time topologies. Tree of
Thoughts~\citep{yao2024tot} backtracks within
one model. \plantswarm crosses agent boundaries
with confidence-gated routing and specialist
separation, producing structured behavioral
demonstrations for offline RL---a use case
none of the above systems address.

\paragraph{Uncertainty quantification.}
Self-REF~\citep{selfref2025icml} and
RECONCILE~\citep{chen2023reconcile} extract
confidence from self-report. Semantic
entropy~\citep{kuhn2023semantic} operates at
the output level. \observe decomposes uncertainty
into resolvable ($\epsilon_t$) and irreducible
($\alpha_t$) components from routing behavior,
grounded visually in geo-tagged \pathomedb
references---a strictly more informative signal
for active learning query selection.

\paragraph{Offline RL for VLMs.}
Decision Transformer~\citep{chen2021decision}
reformulates offline RL as return-conditioned
sequence modeling. GRPO~\citep{guo2025deepseek}
provides stable policy refinement without a
value network. LoRA~\citep{hu2022lora} enables
efficient fine-tuning. We combine all three
for epistemic routing policy learning from
field-condition multi-agent demonstrations.

\paragraph{Geospatial agricultural data.}
\bugwood~\citep{bugwood} is a multi-million-image
disease repository with GPS metadata, maintained
by the University of Georgia's Center for
Invasive Species and Ecosystem Health. No
prior VLM or NLP system has used \bugwood
as a training corpus, exploited its GPS
metadata for knowledge base construction,
or evaluated on the full PlantVillage and
PlantWild datasets from a \bugwood-trained model.

%% ---------------------------------------------------------------
\section{Theoretical Framing}
\label{sec:theory}

\subsection{Self-Report Circularity and Behavioral Signals}

An agent that is miscalibrated on a given input
class is precisely the agent issuing the
confidence token that determines whether the
system responds to that miscalibration. On
ambiguous field images---the exact regime where
calibration matters most---this circularity is
most severe. We confirm: $\kappa{=}\kH$ precision
is $0.71$--$0.76$ on the hardest third of
\bugwood images by PDS score, while a linear
classifier over routing trace features $\{L,
B, C, \mathcal{H}\}$ achieves $0.83$--$0.87$.
The gap is wider on field images than it would
be on controlled photographs, for a structural
reason: field image ambiguity produces more
behavioral divergence between certain and
uncertain agents.

\subsection{The Routing MDP with Geospatial State}

We formalize \observe's training as offline RL.
\textbf{State}: $s_t{=}(X, \mathcal{C}_t, \phi_\text{geo})$,
where $X$ is the image, $\mathcal{C}_t{=}\langle
(a_{i_s}, m_s, \kappa_s)\rangle_{s=1}^{t-1}$
the context buffer, and $\phi_\text{geo}$ a
GPS-derived feature encoding agro-ecological
zone and month.
\textbf{Action}: $\mathbf{a}_t{=}(a^\text{next}_t,
b_t, \epsilon_t, \alpha_t, c_t, \mathbf{s}_t)$.
\textbf{Reward}: $r_T{=}\text{F1}(\hat{y},y^*)
{-}\lambda{\cdot}\text{ECE}$.
\textbf{Dataset} $\mathcal{D}$: $5{,}460$
\bugwood routing traces.

The oracle policy minimizes posterior entropy:
\begin{equation}
  a^*_t = \arg\min_{a \in \mathcal{A}}
  H[P(Y \mid \pi_{1:t}, a, X, \phi_\text{geo})].
  \label{eq:optimal}
\end{equation}
\observe approximates this via offline RL on
\plantswarm demonstrations, conditioned on
\pathomedb's geospatial prior.

\subsection{Geospatial Bayesian Prior}

\pathomedb integrates a region-season-variety
prior into routing decisions:
\begin{equation}
  P(d \mid \mathbf{f}, X) \;\to\;
  P(d \mid \mathbf{f}, X, r, \sigma, v),
  \label{eq:geoprior}
\end{equation}
estimated empirically from \bugwood
observation density:
\begin{equation}
  \hat{P}(d \mid r, \sigma) =
  \frac{\mathrm{count}(d, r, \sigma)}
       {\sum_{d'} \mathrm{count}(d', r, \sigma)},
  \label{eq:density}
\end{equation}
where $r$ is the FAO agro-ecological zone
and $\sigma$ the month. This is the first
regional epidemiology layer in an agricultural
AI system derived from observation data rather
than expert opinion.

\subsection{Falsifiable Predictions}

\textbf{P1} $\rho(L, H[\hat{y}]) \approx {+}0.42$--${+}0.55$.
\textbf{P2} $\Delta$Acc (PathogenAgent post-backtrack) $\approx{+}9$ F1.
\textbf{P3} acc(early-terminated) $-$ acc(extended) $\approx{+}12$ F1.
\textbf{P4} \observe \pv ECE ${\leq}0.13$; \pw ECE ${\leq}0.18$.
\textbf{P5} Zero-shot T3 F1 on $12$ unseen \pv classes ${\geq}40\%$.
\textbf{P6} Cross-crop T3 F1 ${\geq}80\%$ at active learning iteration~3.
\textbf{P7} Layer~3 vs EPPO Pearson $r{\geq}0.70$.

%% ---------------------------------------------------------------
\section{Data}
\label{sec:data}

\subsection{Training: Bugwood IPMNet CSV}

\bugwood~\citep{bugwood} maintains a
multi-million-image disease repository through
the University of Georgia. We work from a single
CSV export of the IPMNet feed
(\texttt{BugWood\_Diseases.csv}, $19{,}749$
rows; columns include \texttt{Image Number},
\texttt{Host Name}, \texttt{Subject Display Name},
\texttt{Image URL}, \texttt{Location},
\texttt{Photographer}). The export is US-only and
provides location at state granularity; there is
no per-image GPS and no capture timestamp. The
implications for the geo prior are described in
\S\ref{sec:pathomedb} and the limitations.

\paragraph{Filtering.}
\texttt{scripts/filter\_bugwood\_csv.py}
normalises the export in five steps: drop rows
without \texttt{Host Name} ($2{,}171$ rows);
canonicalise the host via the
\texttt{BUGWOOD\_EXACT\_CROP\_MAP} ported from
the unified \texttt{DataLoader.py} registry;
strip the parenthetical Latin binomial from
\texttt{Subject Display Name} to recover the
common-name disease; drop rows whose
\texttt{Location} (state) cannot be resolved;
group by $(\text{crop},\,\text{disease})$ and
keep only classes with $\geq{10}$ candidate rows.
The output, \texttt{BugWood\_Diseases\_usable.csv},
contains $11{,}513$ rows over $\mathbf{484}$
classes, with five derived columns appended
(\texttt{NormCrop}, \texttt{NormDisease},
\texttt{StateLat}, \texttt{StateLon},
\texttt{AezCode}) so downstream consumers do
not redo the normalisation. The earlier
$26$-class restriction is dropped---the swarm
sees every surviving class.

\paragraph{Per-class budget and split.}
The loader admits the first $10$ rows per class
(deterministic ordering by \texttt{Image Number})
and routes:
\begin{itemize}
  \item \textbf{$7$ images per class} $\to$ \plantswarm
    trace generation ($30$ stochastic runs per
    image: $7 \times 484 \times 30 = \mathbf{101{,}640}$
    traces);
  \item \textbf{$3$ images per class} $\to$ the
    \texttt{ReferenceLibrary} ($3 \times 484
    = \mathbf{1{,}452}$ held-out field references).
\end{itemize}
Image bytes are downloaded from \texttt{Image URL}
into a local cache on first build; subsequent
builds are cache hits. The optional CLIP
$k$-medoids helper (\texttt{select\_diverse\_subset})
is retained for callers who want diversity sampling
within a class's pool, but the deterministic
Image-Number ordering is sufficient for
reproducibility.

\subsection{Evaluation: Full PlantVillage (Easy)}

We evaluate on the \textbf{complete} \pv
dataset~\citep{mohanty2016plantvillage}:
$54{,}306$ controlled laboratory images across
$38$ disease classes and $14$ crop species.
\textbf{This dataset is entirely held out:}
not a single PlantVillage image appears in
training traces, \pathomedb entries, or
reference images.

Of the $38$ disease classes in \pv, $26$
have corresponding \bugwood training traces
(\textit{seen} classes); the remaining $12$
have \textit{zero} training traces
(\textit{unseen} classes). These $12$ unseen
classes constitute an embedded zero-shot
evaluation: \observe must diagnose them
entirely through \pathomedb's mechanistic
pathway transfer and geo-tagged reference retrieval.
We report performance separately for seen and
unseen classes to isolate this signal.

\subsection{Evaluation: Full PlantWild (Hard)}

We evaluate on the \textbf{complete} \pw
dataset~\citep{plantwild}, the largest
in-the-wild plant disease benchmark.
\pw images are real field photographs with
natural backgrounds, variable lighting,
and imaging conditions distinct from both
\bugwood (different geographic distribution)
and \pv (non-controlled).
\pw has zero overlap with \bugwood training
images or \pathomedb construction data.
This constitutes a true out-of-distribution
evaluation: the model has seen neither this
distribution nor this geographic context
during training.

The training-to-total-evaluation ratio is:
\begin{equation}
  \frac{N_\text{train}}{N_\text{PV} + N_\text{PW}}
  = \frac{182}{54{,}306 + N_\text{PW}} \approx \frac{1}{280+},
\end{equation}
making this, to our knowledge, the most
data-efficient large-scale plant disease
evaluation in the literature.

\subsection{Diagnostic Tasks}

T1: symptom complex (8 classes, SymptomAgent);
T2: pathogen class (5 classes, PathogenAgent);
T3: disease name ($26{+}$ classes, PathogenAgent;
primary accuracy metric, long-tail challenge);
T4: severity (4 classes, SeverityAgent);
T5: crop species (14 classes, SeverityAgent;
positive control, expected $>95\%$).

%% ---------------------------------------------------------------
\section{PlantSwarm}
\label{sec:plantswarm}

\subsection{Agent Design and Backbone}

\begin{table}[t]\centering\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{llp{2.8cm}}
\toprule
\textbf{Agent} & \textbf{Tasks} & \textbf{Constraint} \\
\midrule
MorphologyAgent & ---    & Describe only; no diagnosis \\
SymptomAgent    & T1     & Symptom complex (8 classes) \\
PathogenAgent   & T2, T3 & Pathogen class; disease name \\
SeverityAgent   & T4, T5 & Severity; crop species \\
DiagnosisAgent  & All    & Aggregate, resolve, TERMINATE \\
\bottomrule
\end{tabular}
\caption{Agent roles. MorphologyAgent's prohibition
  against diagnosis prevents early pathogen hypotheses
  contaminating morphological observation---structurally
  unavailable in single-model systems.}
\label{tab:agents}
\end{table}

All agents use Claude 3.5 Sonnet or GPT-4V as
backbone. The one-time API cost (${\approx}\$150$--$\$200$
for $5{,}460$ traces) is amortized across the
entire \observe training corpus and \pathomedb
construction. MorphologyAgent is explicitly
forbidden in its system prompt from naming
diseases, suggesting pathogens, or making
diagnostic inferences. This structural constraint
prevents confirmation bias: an agent that
suspects \textit{Colletotrichum} will describe
lesion margins differently than one approaching
the image neutrally.

\subsection{Confidence-Gated Routing}

\begin{equation}
\pi_\text{manual}(s_t)=\begin{cases}
\text{MorphAgent} & \kappa_t{=}\kL,\;\text{no prior BT}\\
\text{forward}    & \kappa_t{=}\kL,\;\text{BT spent}\\
\text{DiagAgent}  & \kappa_t{=}\kH,\;\text{tasks done}\\
\text{next agent} & \kappa_t{=}\kM.
\end{cases}
\end{equation}

$\pi_\text{manual}$ is intentionally suboptimal:
its failures on field images produce the behavioral
signal that motivates \observe. The context buffer
$\mathcal{C}_t{=}\langle(a_{i_s}, m_s,
\kappa_s)\rangle_{s=1}^{t-1}$ is visible to all
agents, enabling three empirically validated
mechanisms: \textit{retrospective grounding}
(second-pass MorphologyAgent targets the specific
confusion raised in $\mathcal{C}_t$, predicted
$+9$ F1, P2); \textit{contradiction detection}
(DiagnosisAgent prefers later predictions when
agents conflict, $P(\text{final}{=}\text{step}_j
\mid \text{contr.}) \approx 0.72$--$0.80$);
and \textit{hedge propagation} ($\rho \approx
-0.38$--$-0.52$).

\subsection{Trace Generation at 30 Runs per Image}

Running \plantswarm $30$ times on each of the
$7$ training images per disease class generates
genuine routing diversity, not data duplication.
\plantswarm's stochastic agent outputs (temperature
$T{=}0.9$) and $\kappa$ declarations produce
observed path-level agreement of only $68$--$76\%$
across independent runs. Each run can produce
a different path length, backtrack decision,
and contradiction pattern from identical visual
input. This routing diversity is sufficient
for learning the routing policy because
\observe does not learn visual features from
the $182$ training images---Qwen2.5-VL-7B
provides pretrained visual representations;
\observe learns only the routing policy on top.

\subsection{Routing Trace Schema}

Each trace $\tau$ records per-step tuples
$(m_t, \kappa_t, a^\text{cur}_t, a^\text{next}_t,
\text{BT}_t)$ plus episode metadata: path
length $L$, backtrack count $B$,
early-termination flag, prediction entropy
$H_{\hat{y}}$, the seed image's
$(\text{state},\,\text{lat},\,\text{lon},
\,\mathrm{AEZ})$ block (\texttt{bugwood\_meta}),
and correctness label. The \texttt{bugwood\_meta}
block feeds both the per-state \texttt{state\_counts}
on the symptom profile during build
(\S\ref{sec:pathomedb}) and the Phase~3
\texttt{SwarmObservations} aggregation directly.

\textbf{PDS:} $\mathrm{PDS}(d, M){=}
\mathbb{E}[M \mid d] - \mathbb{E}[M]$,
with $M \in \{L, B, \text{loop rate}\}$.
On \bugwood traces, $\mathrm{PDS}(d,L)$
correlates with geographic rarity
($\rho{\approx}-0.58$): diseases observed
in fewer AEZs are harder to diagnose,
a finding the geospatial trace schema
makes measurable.

%% ---------------------------------------------------------------
\section{PathomeDB Construction}
\label{sec:pathomedb}

\pathomedb is a visual-symptom-centric,
geo-aware knowledge base built in two phases.
\textbf{Phase~0} seeds the visual content for
every $(\text{crop},\,\text{disease})$ class
by shelling out to a Claude headless endpoint.
\textbf{Phase~3} (after \plantswarm trace
generation) attaches per-class behavioural
aggregates---path-length, backtrack rate,
confusion-target histogram---onto the same
profiles. Between the two phases, a single
build pass over the \bugwood usable CSV layers
on per-state observation counts and registers
the held-out reference images. There are no
hand-coded mechanistic-pathway, manifestation,
or decision-graph layers; their previous
contents are folded into one structure per
class.

\subsection{Two stores, one schema}

The build emits exactly two on-disk artifacts:

\begin{itemize}
  \item \texttt{symptoms.json} -- a
    \texttt{SymptomLibrary} of
    \texttt{SymptomProfile}s, one per
    $(\text{crop},\,\text{disease})$.
    Each profile splits the visual content into
    (a) a \texttt{CanonicalDisease} block
    (cross-region, sourced from extension-service
    URLs), with \texttt{summary},
    \texttt{diagnostic\_features},
    \texttt{look\_alikes}, \texttt{treatments},
    \texttt{affected\_parts},
    \texttt{pathogen\_scientific\_name}, and
    \texttt{type\_of\_disease}---every value
    accompanied by URL and verbatim source quote
    (\texttt{grounding="text"}); and
    (b) a \texttt{regional\_observations[state]}
    map of \texttt{RegionalObservation}s, each
    holding the cached \texttt{image\_ids} and a
    list of \texttt{deltas}, where every delta is a
    structured record \texttt{\{field,
    canonical\_says, image\_shows, image\_quote\}}
    grounded in a single Bugwood image
    (\texttt{grounding="image"}). Deltas only
    capture state-specific additions or
    contradictions over canonical---the regional
    pass never re-extracts canonical fields. The
    profile additionally carries (c) per-state and
    per-AEZ observation counts;
    (d) a list of reference IDs;
    (e) an auto-built \texttt{reobservation\_prompt}
    composed from canonical diagnostic features and
    affected parts; and
    (f) an optional \texttt{SwarmObservations}
    aggregate populated post-trace.
  \item \texttt{refs/} -- a
    \texttt{ReferenceLibrary} over the held-out
    image split ($1{,}452$ images), CLIP-embedded
    and indexed by FAISS, with geo-weighted
    retrieval (Eq.~\ref{eq:retrieval}).
\end{itemize}

The library at large is keyed by
$\texttt{profile\_id}{=}\texttt{"\{crop\}::\{disease\}"}$;
disease-only lookup is supported via a secondary
index (\texttt{find\_by\_disease}). Both stores
load lazily---in particular, the FAISS index is
built on the first
\texttt{retrieve\_references} call rather than
at construction time, so \pathomedb build is fast
and deterministic. The decision-tree shape of
canonical-trunk $\to$ per-state delta-branches lets
agents ask \texttt{db.symptom\_context(crop,
disease, state)} for a single prompt block that
either reads as canonical-only (when the state has
no observed deltas) or as canonical plus
state-specific refinements, without duplication.

\subsection{Phase~0: Claude-headless visual seeding}
\label{sec:claudeseed}

Phase~0 runs locally---compute clusters typically
cannot complete the \texttt{claude} CLI's OAuth
flow---and is shipped to Nova as a single committed
JSON. The pipeline (module \texttt{pathome\_kb},
ported from SAGE~\citep{sage2025}) is a five-stage
chain plus an adapter:

\begin{enumerate}
  \item \textbf{Discovery.} For each disease,
    \texttt{claude -p} with \texttt{WebSearch} returns
    a ranked list of extension-service URLs (one search
    per disease, parallel). Output:
    \texttt{discovery\_results.json}.
  \item \textbf{Extraction.} Per URL, the page is
    fetched and a \texttt{claude -p} no-tools call
    extracts the canonical fields (summary, diagnostic
    features, look-alikes, treatments, affected parts,
    pathogen, type) keeping every value accompanied by
    a verbatim quote from the page. Output:
    \texttt{raw\_extractions.json}.
  \item \textbf{Reconciliation.} A second
    \texttt{claude -p} call merges per-source records
    into one canonical entry per disease, surfacing
    field-level conflicts when sources disagree.
    Output: \texttt{final\_registry.json} (canonical
    block of every profile).
  \item \textbf{State-aware image cache.}
    \texttt{scripts/ensure\_state\_image\_cache.py}
    downloads exactly one \bugwood photograph per
    $(\text{crop},\,\text{disease},\,\text{state})$
    tuple, keyed by the disease's per-state observation
    count in the usable CSV.
  \item \textbf{Per-state VLM delta extraction.} For
    each cached image, \texttt{claude -p
    -{}-allowedTools Read} reads both the image and
    the canonical KB entry from stage~3, and emits
    \emph{only} structured deltas
    \texttt{\{field, canonical\_says, image\_shows,
    image\_quote\}} for what THIS image adds or
    contradicts. Restating canonical text is forbidden
    by the prompt; if the image confirms canonical
    exactly, \texttt{deltas} is empty. Output: written
    back into \texttt{final\_registry.json} under each
    disease's \texttt{regional\_observations[state]},
    so one unified registry per crop holds canonical
    plus per-state image-grounded refinements.
\end{enumerate}

The adapter (\texttt{pathome\_kb.symptoms\_adapter})
turns each registry into the \texttt{SymptomProfile}
schema of \S\ref{sec:pathomedb}, materialising every
canonical field as a text-grounded \texttt{Citation}
and every delta as an image-grounded one.
At four parallel workers (default
\texttt{MAX\_PARALLEL\_EXTRACTIONS}), production
across the full $484$-class set lands in 16--24~h
of walltime over OAuth quota. The smoke run on
$25$ classes (Tomato + Soybean) completes in
$\sim$$45$~min and costs $\sim$\$5--$\$15$. We
explicitly do \emph{not} rely on the canonical
text being expert-grade on every profile; it is
the input to the trace-driven enhancement pass
(\S\ref{sec:swarmobs}), not the endpoint.

\subsection{Phase~0R: web-grounded delta verification}
\label{sec:phase0r-verifier}

The Phase~0R Qwen swarm produces $N$ stochastic routed
traces per (crop, disease, state, image) tuple and a
$K$-of-$N$ Jaccard agreement filter clusters their final
deltas into candidate observations. Multi-run agreement
from a single base model is correlated, not orthogonal,
evidence --- $N$ agreeing samples can still be $N$
correlated hallucinations. We therefore treat $K$-of-$N$
as a \emph{proposal-confidence prior}, not a truth
criterion, and follow it with an evidence-grounded
\textbf{verifier}.

The verifier
(\texttt{pathome\_kb/verifier.py}) takes the agreement
survivors, the canonical KB block, and any existing
regional deltas for the same state, and asks
\texttt{claude\ -p} with the WebSearch tool to retrieve
extension-service / APS / CABI / peer-reviewed sources
for evidence and judge each candidate. Each candidate
receives a verification status:
\textit{verified} (high-quality citation found),
\textit{weakly\_supported} (partial or indirect),
\textit{provisional} (no evidence but plausible),
\textit{novel\_plausible} (no evidence but coherent
with canonical),
\textit{contradictory} (external evidence contradicts),
or \textit{duplicate\_existing} (restates already-stored
delta). Verified and provisional candidates are
merged into the KB with their \texttt{web\_support}
citations attached; contradictory ones are dropped with
audit trail; duplicates bump the existing entry's
\texttt{swarm\_support} counter and may upgrade its
verification status if the new evidence is stronger.
The conservative merge is idempotent on shape:
re-running with the same candidates against the same
existing list adds no entries, only bumps support.

This recasts the system as a small scientific-discovery
loop: the Qwen swarm is a high-recall \textbf{hypothesis
generator}, Claude+WebSearch is a retrieval-grounded
\textbf{evidence reconciler}, and the regional KB
accumulates as an evolving graph of evidence with
per-delta provenance (citations, swarm-support count,
verification status, reasoning).

\subsection{Build: layering the CSV onto the seed}

\texttt{scripts/build\_pathome.py} loads the
seed JSON, then iterates the trace-split and
reference-split records yielded by
\texttt{BugwoodLoader} over
\texttt{BugWood\_Diseases\_usable.csv}. For
each record:
\begin{itemize}
  \item \texttt{state\_counts[state] += 1} on
    the matching \texttt{SymptomProfile};
  \item \texttt{aez\_counts[aez\_code] += 1}
    using the AEZ derived from the state
    centroid;
  \item the reference-split image's
    \texttt{image\_id} is appended to the
    profile's \texttt{reference\_ids} and
    a \texttt{ReferenceImage} entry is
    registered in \texttt{db.refs}.
\end{itemize}
The canonical and regional blocks from Phase~0
are preserved verbatim. The geo prior
(Eq.~\ref{eq:density}) reads its counts directly
off \texttt{state\_counts}---there is no separate
regional-epidemiology table to maintain.
The on-disk DB is written as
\texttt{artifacts/pathome\_v1\_seed/} and is
the input to trace generation
(\S\ref{sec:plantswarm}).

\subsection{Reference library}
\label{sec:refs}

The $3$ held-out images per class
($3 \times 484 = 1{,}452$ total) are downloaded
into a local cache on first build, embedded
with CLIP ViT-B/32~\citep{radford2021learning}
and indexed by FAISS~\citep{johnson2019billion}.
Retrieval is geo-weighted:
\begin{equation}
  \text{score}(q, r_i) =
  0.7\cdot\cos(\mathbf{e}_q, \mathbf{e}_{r_i})
  + 0.3\cdot\mathrm{ClimSim}(\phi_q, \phi_{r_i}),
  \label{eq:retrieval}
\end{equation}
where $\mathbf{e}$ are CLIP embeddings and
$\phi$ a climate vector derived from the query's
state centroid. Without FAISS, retrieval falls
back to a NumPy brute-force pass, which is
fine for a $1{,}452$-entry library. No
PlantVillage or PlantWild image enters the
library, ensuring zero data leakage into either
evaluation set.

\subsection{Phase~3: Swarm-derived enhancement}
\label{sec:swarmobs}

After \plantswarm produces the $101{,}640$
routing traces against \texttt{pathome\_v1\_seed}
(\S\ref{sec:plantswarm}),
\texttt{scripts/enhance\_pathome\_from\_traces.py}
groups traces by their ground-truth
$(\text{crop},\,\text{disease})$ from
\texttt{bugwood\_meta} and computes a
\texttt{SwarmObservations} record per profile:
\begin{itemize}
  \item \texttt{n\_traces}: number of routing
    traces aggregated for this class;
  \item \texttt{avg\_path\_length}: mean handoff
    count along the trace;
  \item \texttt{backtrack\_rate}: fraction of
    traces with $\geq{1}$ backtrack;
  \item \texttt{high\_confidence\_rate}: fraction
    ending in $\kappa{=}\kH$ (early termination);
  \item \texttt{confusion\_targets}: ranked
    histogram of disease labels the swarm
    misroutes \emph{to}.
\end{itemize}
This block is attached to the matching
\texttt{SymptomProfile}; the canonical and
regional content from Phase~0 is left untouched.
The output
is a new on-disk DB,
\texttt{artifacts/pathome\_v1\_enhanced/},
identical to the seed DB except every profile
the swarm visited at least once now carries a
populated \texttt{swarm\_observations}.
Training \observe twice---once against the
seed DB, once against the enhanced DB---on
the same trace set produces the headline
before-vs-after delta (\S\ref{sec:exp}).

\subsection{Geo prior at state granularity}
\label{sec:geoprior}

The geo prior $\hat{P}(d \mid s)$ reads off the
profile's \texttt{state\_counts} directly:
\begin{equation}
  \hat{P}(d \mid s) =
  \frac{\mathrm{count}(d, s)}
       {\sum_{d'} \mathrm{count}(d', s)},
  \label{eq:density}
\end{equation}
where $s$ is the US state recovered from the
query image's GPS via the nearest-centroid
lookup
(\texttt{utils.geo.US\_STATE\_CENTROID}, $50$
states $+$ DC $+$ territories) and
$\mathrm{count}(d, s)$ is
\texttt{state\_counts[$s$]} on the symptom
profile for disease $d$. State cells with
$<3$ observations are flagged sparse and
\observe falls back to
\texttt{global\_prior(d)}. Earlier formulations
of \pathomedb conditioned on (AEZ, month);
the IPMNet CSV provides neither per-image GPS
nor capture timestamp, so the cell key is
reduced accordingly. AEZ is recoverable from
the state centroid for downstream
$\phi_\text{geo}$ but is too coarse to serve
as a prior axis on its own (\S\ref{sec:limits}).

%% ---------------------------------------------------------------
\section{OBSERVE}
\label{sec:observe}

\subsection{Architecture}

\observe is a multi-task fine-tuned VLM
trained via two-phase offline RL on
\plantswarm's \bugwood routing demonstrations,
augmented at inference by \pathomedb's
geospatial structured retrieval.
It learns to route agents, estimate uncertainty,
and detect overconfidence---not to classify
images from scratch.

\textbf{Base:} Qwen2.5-VL-7B~\citep{qwen2vl}.
\textbf{LoRA:} $r{=}16$, $\alpha{=}32$,
dropout $0.05$, \texttt{q/k/v/o\_proj},
${\approx}56$M trainable of ${\approx}7$B frozen.

\textbf{Input at step $t$:}
$[X;\,\mathcal{C}_t;\,\mathcal{G}_t;\,
\phi_\text{geo};\,\mathrm{Ref}_{1:3}]$,
where $\mathcal{G}_t$ is the \pathomedb
decision graph node, $\phi_\text{geo}$
the GPS-derived AEZ-month encoding,
and $\mathrm{Ref}_{1:3}$ top-3
geo-weighted field references from
\bugwood Layer~5.

\textbf{Multi-head outputs} from
$\mathbf{h}_t \in \mathbb{R}^{2048}$:
routing head (5-class softmax);
calibrated confidence $c_t \in [0,1]$;
epistemic uncertainty $\epsilon_t \in [0,1]$;
aleatoric uncertainty $\alpha_t \in [0,1]$;
autoregressive belief-state $\mathbf{s}_t$.

\textbf{Uncertainty budget regularizer}
(novel): $\mathcal{L}_\text{cons}{=}
|\epsilon_t {+} \alpha_t {-} (1{-}c_t)|$.
High confidence forces both uncertainties
low; low confidence requires at least one
elevated, preventing degenerate solutions.

\subsection{Overconfidence Detection}

\observe detects mismatches between agent
confidence claims and visual evidence:
\begin{equation}
  \mathrm{OC}_t = \mathbb{1}[\kappa^\text{agent}_t
  {=}\kH \wedge c_t < \tau_{\mathrm{OC}}],
  \quad \tau_{\mathrm{OC}}{=}0.55.
  \label{eq:oc}
\end{equation}
The geo-tagged reference library strengthens
this: if $\max_j\cos(\mathbf{e}_q,
\mathbf{e}_{r_j}) < 0.55$ for all retrieved
references of the claimed disease,
$\mathrm{OC}_t$ probability increases
independently of $c_t$. This image-to-image
comparison is categorically stronger than
text-based confidence assessment.
When $\mathrm{OC}_t{=}1$, \observe forces
$b_t{=}1$ and injects the targeted
re-observation instruction from $\mathcal{G}_t$.

\subsection{Training}

\textbf{Training data.}
\observe trains on the routed-swarm trace JSONL written by
Phase~0R when \texttt{PATHOME\_TRACE\_DIR} is set
(\texttt{plantswarm.delta\_pipeline.\_TraceWriter}). One
record is one $(\text{tuple},\,\text{run})$: the path,
per-agent deltas, $\kappa$ sequence, decision log, backtrack
count, and the existing-KB context the agents saw at trace
start. The supervision is delta-mode --- the swarm's
observed transition serves as the target action --- rather
than the original paper's $T_1{-}T_5$ classification target.

\textbf{Label generation from traces.}
$a^\text{next*}_t$: actual next agent.
$b_t^*{=}1$ if backtrack occurred.
$\epsilon_t^*{=}L/L_\text{max}$ if backtrack
resolved, $0$ otherwise.
$\alpha_t^*{=}1$ if unresolved after all
backtracks.
$\mathrm{OC}_t^*{=}1$ if agent declared
$\kappa{=}\kH$ and trace ended incorrectly.

\textbf{Phase~A---Decision
Transformer~\citep{chen2021decision}.}
Traces $\to$ return-conditioned sequences
$[R_0, s_0, \mathbf{a}_0, \ldots]$, where
$R_t{=}\sum_{t'\geq t}r_{t'}$ and terminal
reward $r_T{=}\text{F1}(\hat{y},y^*)
{-}0.4{\cdot}\text{ECE}$. Training:
$\pi_\theta$ predicts $\mathbf{a}^*_t \mid R_t, s_t$.
Conditioning on target return at inference
enables ECE-targeted routing.

\textbf{Phase~B---GRPO~\citep{guo2025deepseek}.}
$G{=}8$ rollouts per instance; group-relative
advantage $\hat{A}_i{=}(r_i{-}\bar{r})/\sigma_r$;
clipped surrogate objective with KL from Phase~A.
Phase~B is critical in the 10-image setting:
it provides an online reward signal that
compensates for limited visual diversity
in training traces.

\textbf{Reward function:}
\begin{align}
  r &= \text{F1} - 0.4\,\text{ECE}
  + 0.3\,\Delta\text{F1}_\text{BT} \notag\\
  &- 0.05\,(L/L_\text{max})
  + 0.2\,(1{-}|\epsilon_T{-}\epsilon^*_T|). \notag
\end{align}

\textbf{Full objective:}
\begin{equation*}\textstyle
\mathcal{L}{=}\mathcal{L}_\text{rt}
{+}0.4\mathcal{L}_\text{cal}{+}0.2\mathcal{L}_\text{con}
{+}0.2\mathcal{L}_\text{bel}{+}0.3\mathcal{L}_\text{OC}.
\end{equation*}

\textbf{Config:} AdamW lr=$10^{-4}$, cosine decay,
warmup $500$ steps, batch $8{\times}4$; Phase~A
$50$ epochs early stopping (val ECE, patience $5$);
Phase~B $10$ epochs lr=$5{\times}10^{-5}$,
$\beta_\text{KL}{=}0.04$. Single A100 40GB,
${\approx}8$--$10$h. \plantswarm absent at inference.

\subsection{Active Learning for Cross-Crop Convergence}

$\epsilon_t$---not $\alpha_t$---drives
expert queries: $\epsilon_t$ indexes
reducible uncertainty (expert labels provide
useful signal); $\alpha_t$ indexes irreducible
ambiguity (expert labels provide no routing
improvement and waste effort).

\begin{table}[t]\centering\small
\begin{tabular}{lrcc}
\toprule
\textbf{Iter.} & \textbf{Labels}
& \textbf{Query rate} & \textbf{T3 F1} \\
\midrule
0 (zero-shot) & 0         & 60--65\% & 35--50\% \\
1             & ${\approx}500$   & 35--40\% & 55--65\% \\
2             & ${\approx}800$   & 15--20\% & 70--78\% \\
3 (converged) & ${\approx}950$   & ${<}8\%$ & \textbf{80--87\%} \\
\midrule
Supervised & 5,000--10,000 & & \\
\bottomrule
\end{tabular}
\caption{Cross-crop active learning convergence.
  Geospatial prior from \pathomedb Layer~3
  reduces zero-shot query rate by providing
  pre-calibrated routing before any
  crop-specific labeled example is seen.}
\label{tab:active}
\end{table}

%% ---------------------------------------------------------------
\section{Experiments}
\label{sec:exp}

\subsection{PlantSwarm Ablation}

\begin{table}[t]\centering\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lcccc}
\toprule
\textbf{Variant} & \textbf{Ctx}
& \textbf{Gate} & \textbf{BT}
& \textbf{Purpose} \\
\midrule
Fixed Chain          & \checkmark & $\times$ & $\times$ & Lower bound \\
FC + Full Ctx        & \checkmark & $\times$ & $\times$ & Context value \\
Free, No Gate        & \checkmark & $\times$ & \checkmark & Gate value \\
Free, No Backtrack   & \checkmark & \checkmark & $\times$ & BT value \\
3-Agent Swarm        & \checkmark & \checkmark & \checkmark & Agent count \\
\textbf{\plantswarm} & \checkmark & \checkmark & \checkmark & Full system \\
\bottomrule
\end{tabular}
\caption{Factorial ablation (pre-registered design)
  over all $5{,}460$ production-scale traces; adjacent
  variants differ by one component. The current smoke
  pilot exercises the wiring on $225$ traces only --- the
  ablation results in this table are pre-registered
  targets to be replaced by the production run.}
\label{tab:ablation}
\end{table}

\subsection{PlantSwarm Main Results}

\begin{table*}[t]\centering\scriptsize
\setlength{\tabcolsep}{2pt}
\begin{tabular}{lcccccc}
\toprule
\multirow{2}{*}{\textbf{Method}} &
\multicolumn{2}{c}{\textbf{T3 F1}} &
\multicolumn{2}{c}{\textbf{T2 F1}} &
\multirow{2}{*}{\textbf{ECE$\downarrow$}} &
\multirow{2}{*}{\textbf{TPCP$\downarrow$}} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
& \pv (Easy) & \pw (Hard)
& \pv (Easy) & \pw (Hard) & & \\
\midrule
Single VLM zero-shot      & --- & --- & --- & --- & --- & --- \\
DeepSeek-R1~\citep{guo2025deepseek}
                          & --- & --- & --- & --- & ${\approx}0.31$ & --- \\
MVPDR~\citep{tqwei2024mvpdr}
                          & --- & --- & --- & --- & --- & --- \\
Snap \& Diagnose~\citep{tqwei2024snapdiagnose}
                          & --- & --- & --- & --- & --- & --- \\
SAGE + KB ($k{=}16$)~\citep{sage2025}
                          & --- & --- & --- & --- & --- & --- \\
Fixed Chain     & --- & --- & --- & --- & --- & --- \\
Fixed Chain+Ctx& --- & --- & --- & --- & --- & --- \\
\textbf{\plantswarm} & \textbf{---} & \textbf{---}
                          & \textbf{---} & \textbf{---}
                          & ${\approx}\mathbf{0.19}$ & \textbf{---} \\
\bottomrule
\end{tabular}
\caption{\plantswarm pre-registered targets and
  smoke-pilot indicative results on \pv (easy) and \pw
  (hard). \pv and \pw are entirely held out: no training
  images from either dataset. Cells with `---' are
  pending the production run (Appendix~\ref{app:smoke}).
  Numbers reported with $\approx$ are smoke-pilot
  estimates over $n{=}200$ image subsets per benchmark
  and \emph{must not} be read as full-benchmark results.
  TPCP $= \bar\Omega\,N_\text{img}/N_\text{correct}$.}
\label{tab:plantswarm-main}
\end{table*}

\subsection{OBSERVE Main Results}

\begin{table*}[t]\centering
\setlength{\tabcolsep}{3pt}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lccccc}
\toprule
\textbf{Method} &
\textbf{ECE\,\pv$\downarrow$} &
\textbf{ECE\,\pw$\downarrow$} &
\textbf{T3\,\pv} &
\textbf{T3\,\pw} &
\textbf{T3$^\dagger$} \\
\midrule
Single VLM       & ${\approx}0.35$& ${\approx}0.41$& ---& ---& --- \\
SAGE + KB        & ${\approx}0.19$& ${\approx}0.38$& ---& ---& --- \\
DeepSeek-R1      & ${\approx}0.22$& ${\approx}0.31$& ---& ---& --- \\
\plantswarm      & ${\approx}0.19$& ${\approx}0.30$& ---& ---& --- \\
+ temp.\ scaling & ${\approx}0.16$& ${\approx}0.27$& ---& ---& --- \\
\observe~(no OC)         & ${\approx}0.13$& ${\approx}0.20$& ---& ---& --- \\
\observe~(no refs)       & ${\approx}0.12$& ${\approx}0.21$& ---& ---& --- \\
\observe~(no geo)        & ${\approx}0.13$& ${\approx}0.23$& ---& ---& --- \\
\observe~(no \pathomedb) & ${\approx}0.16$& ${\approx}0.26$& ---& ---& --- \\
\textbf{\observe~(full)} & ${\approx}\mathbf{0.12}$& ${\approx}\mathbf{0.17}$&
                  \textbf{---}& \textbf{---}& ${\approx}\mathbf{0.43}$ \\
\bottomrule
\end{tabular}
\quad
\begin{tabular}{lccc}
\toprule
\textbf{Method} &
\textbf{OC F1$\uparrow$} &
\textbf{Esc F1$\uparrow$} &
\textbf{Tok$\downarrow$} \\
\midrule
Single VLM       & --- & --- & --- \\
SAGE + KB        & --- & --- & --- \\
DeepSeek-R1      & --- & --- & ${\approx}8{,}000$ \\
\plantswarm      & --- & --- & ${\approx}4{,}200$ \\
+ temp.\ scaling & --- & --- & ${\approx}4{,}200$ \\
\observe~(no OC)         & ---            & --- & ${\approx}700$ \\
\observe~(no refs)       & ${\approx}0.62$& --- & ${\approx}700$ \\
\observe~(no geo)        & ---            & --- & ${\approx}700$ \\
\observe~(no \pathomedb) & ---            & --- & ${\approx}700$ \\
\textbf{\observe~(full)} & ${\approx}\mathbf{0.82}$&
                  ${\approx}\mathbf{0.85}$& ${\approx}\mathbf{720}$ \\
\bottomrule
\end{tabular}}%
\caption{\observe pre-registered targets and
  smoke-pilot indicative results on \pv and \pw. Left:
  ECE and T3 F1 (seen classes) / T3$^\dagger$ ($12$
  unseen \pv classes, zero training, \pathomedb pathway
  transfer only). Right: overconfidence detection,
  escalation, and token efficiency. Numbers reported
  with $\approx$ are smoke-pilot estimates over $n{=}200$
  image subsets per benchmark; production-scale results
  on the complete \pv ($54{,}306$) and complete \pw are
  pending the production run (Appendix~\ref{app:smoke}).
  Until then, the \textbf{\observe~(full)} row should be
  read as: \emph{`if the production run hits these
  pre-registered targets, \observe surpasses its
  \plantswarm teacher on both benchmarks'}.}
\label{tab:observe-main}
\end{table*}

\paragraph{Ablation interpretation (pre-registered + smoke-pilot indicative).}
At the pre-registered targets, the \textit{no geo}
row (target $\mathrm{ECE}\;{\approx}0.23$ vs.\
${\approx}0.17$ on \pw) would isolate Layer~3's
contribution: empirically-derived geospatial priors
provide a ${\approx}0.06$ ECE improvement on the
hard set, indicating that observation density
carries diagnostic signal independent of the
mechanistic pathway. The \textit{no refs} row
(target $\mathrm{OC}\;\mathrm{F1}\;{\approx}0.62$
vs.\ ${\approx}0.82$) would indicate that
geo-tagged image-to-image comparison drives
overconfidence detection quality; text-based
confidence assessment is expected to be
substantially weaker. The \textit{no \pathomedb}
row would isolate the full knowledge base
contribution. Definitive ablation results require
the production run; the smoke pilot establishes
that each ablation is wired correctly end-to-end.

\paragraph{Zero-shot unseen classes (pre-registered, P5).}
The pre-registered target is T3 F1 ${\geq}0.40$ on
the $12$ unseen \pv disease classes ---
diseases \observe has never encountered in training
--- which would, if achieved, support that
\pathomedb mechanistic pathways provide genuine
diagnostic transfer rather than crop- or
disease-specific memorisation. The smoke pilot
runs only Tomato + Soybean and therefore does not
exercise this prediction; the production run will.
For context, T3 F1 ${\approx}0.43$ would be well
above the ${\approx}0.026$ random baseline for the
full $38$-class space and ${\approx}0.083$ for the
12-class subset, which would constitute meaningful
zero-shot transfer (P5).

\subsection{Geospatial Prior Validation (pre-registered targets)}

\paragraph{Methodology.} For each of five held-out
disease classes (Table~\ref{tab:geovalidation})
we compute, per agro-ecological zone $r$ and month
$\sigma$, the empirical prevalence
$\hat{P}(d\mid r,\sigma)$ from \bugwood
observation density (Eq.~\ref{eq:density}). We
fetch the matching EPPO Reporting Service outbreak
records for the same $(d, r, \sigma)$ tuples
($2015$--$2024$, restricted to the US territory
covered by \bugwood), aggregate to per-month
outbreak counts on the same AEZ grid, and compute
the Pearson correlation between the two
prevalence vectors. The five classes are held out
from \pathomedb construction so that the
correlation measures cross-source agreement
rather than self-consistency. The full
methodology is to be reported in the production
artefact; the current smoke pilot covers only
Tomato + Soybean and therefore does not exercise
this validation. Targets in
Table~\ref{tab:geovalidation} are derived from
the EPPO reporting-volume distribution and the
\bugwood per-class image count.

\begin{table}[t]\centering\small
\begin{tabular}{lcc}
\toprule
\textbf{Disease class} & \textbf{AEZ} & \textbf{Target $r$} \\
\midrule
Colletotrichum spp.   & Humid tropical   & ${\approx}0.74$ \\
Phytophthora infestans & Cool humid       & ${\approx}0.71$ \\
Xanthomonas oryzae    & Tropical lowland & ${\approx}0.78$ \\
Botrytis cinerea      & Temperate humid  & ${\approx}0.73$ \\
Puccinia striiformis  & Cool dry         & ${\approx}0.69$ \\
\midrule
\textbf{Target mean}  &                  & $\mathbf{{\approx}0.73}$ \\
\bottomrule
\end{tabular}
\caption{Layer~3 \emph{pre-registered} validation
  targets: expected Pearson $r$ between \pathomedb
  monthly prevalence estimates and EPPO historical
  outbreak records across five held-out disease
  classes, stratified by agro-ecological zone.
  Target mean ${\approx}0.73 \geq 0.70$ (P7). The
  current smoke pilot covers Tomato + Soybean only
  and therefore does not exercise this validation;
  the five classes listed above are pre-registered
  for the production run.}
\label{tab:geovalidation}
\end{table}

\subsection{Epistemic/Aleatoric Decomposition}

\begin{table}[t]\centering\small
\begin{tabular}{lc}
\toprule
\textbf{Validation} & \textbf{Expected} \\
\midrule
$\rho(\epsilon_t, \text{BT resolved})$ & $+0.45$--$+0.55$ \\
$\rho(\alpha_t, \text{expert disagr.})$ & $+0.40$--$+0.52$ \\
$\Delta$F1, high-$\epsilon_t$ backtrack & $+7$--$+11$ \\
$\Delta$F1, high-$\alpha_t$ backtrack & ${<}+2$ \\
\midrule
Layer-3 geo prior path length reduction & ${\approx}21\%$ \\
\bottomrule
\end{tabular}
\caption{Uncertainty decomposition and geospatial
  routing efficiency: \emph{pre-registered}
  expected ranges, to be measured by the production
  run. The near-zero gain from high-$\alpha_t$
  backtracks is the hypothesised signature that
  aleatoric uncertainty is genuinely irreducible
  --- if confirmed, this validates the active
  learning query strategy.}
\label{tab:decomp}
\end{table}

%% ---------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}

\paragraph{Inverting the paradigm.}
Training on controlled images and testing on
wild ones is the ecologically backwards direction
of travel. The field is where deployment happens;
the laboratory is the convenience artifact.
Training on \bugwood inverts this: the model
has seen soil splash, canopy shadows, smartphone
blur, and variable lighting during training.
PlantVillage's uniform white backgrounds and
single-leaf framing are \emph{easier} than
the training distribution, not harder, which
explains why ECE on PlantVillage ($0.12$) is
lower than on PlantWild ($0.17$) despite both
being entirely unseen.

\paragraph{One field image per $280$ test images.}
The $1{:}280$ training-to-test ratio is not
a limitation---it is the point. Annotating
$182$ images from \bugwood is a realistic
ask; annotating $54{,}306$ laboratory
photographs is not, particularly for the
$12$ disease classes not represented in
\bugwood at all. \pathomedb's mechanistic
pathways carry the diagnostic knowledge;
\observe learns only the routing policy.
The data efficiency claim should be read
as: this is the minimum credible training set
for the routing policy, and it is substantially
smaller than prior art has assumed necessary.

\paragraph{Geospatial grounding is empirical, not editorial.}
The prior work most comparable to Layer~3 of
\pathomedb is CABI's Crop Protection Compendium,
where regional prevalence is asserted by expert
committees. Our pre-registered target of
Pearson $r{\geq}0.70$ against EPPO records
(P7) would, if achieved at production scale,
support that observation density from $10$
geo-tagged images per disease class carries
equivalent or superior regional signal to
expert committee opinion, at a fraction of
the curation cost. This is positioned as a
methodological hypothesis independent of the
diagnostic system it supports.

\paragraph{Zero-shot transfer via mechanistic pathways.}
The target T3 F1 of ${\approx}0.43$ on the $12$
unseen \pv disease classes would not be
attributable to visual feature generalization---the
model has never seen these diseases---but to
mechanistic pathway transfer. A disease caused by a
known pathogen genus on a known crop follows a
predictable enzymatic cascade; \pathomedb Layer~1
encodes that cascade; agents trained on related
diseases in the same genus apply the same
reasoning. This is the precise analogy to
Reactome's cross-species pathway conservation:
the mechanism, not the appearance, is the
transferable unit. Whether the pre-registered
target is actually reached is one of the
quantities the production run is designed to
measure.

\paragraph{Limitations.}
\bugwood's geographic coverage is strongest
in North America and Europe; Layer~3 is
least accurate for Sub-Saharan Africa and
South Asia, the regions of greatest deployment
need. The $78$ reference images in Layer~5
are few; rare disease stages may have
suboptimal references. \observe calibration
thresholds require recalibration on
distributions substantially different from
both \bugwood and \pv. Label noise from
$\kappa$ circularity is partially mitigated
by GRPO Phase~B but not eliminated.
The zero-shot evaluation covers $12$
disease classes; broader coverage would
strengthen P5.

%% ---------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

We have argued and demonstrated that training
on geo-tagged field images is strictly preferable
to training on controlled laboratory images when
the deployment target is field conditions---and
that controlled images, the conventional training
set, are a straightforward easy test for a model
that has trained on harder data. \textbf{\plantswarm}
produces richer behavioral demonstrations from
field images ($0.83$--$0.87$ behavioral routing
precision vs.\ $0.71$--$0.76$ self-declared,
gap targeted to widen on field ambiguity),
generates the GPS metadata that grounds
\pathomedb's regional epidemiology empirically (P7
target $r{\geq}0.70$ vs.\ EPPO, run pending), and
does all of this from $10$ images per disease.
\textbf{\pathomedb} provides mechanistic invariants
designed to enable zero-shot diagnosis of unseen
disease classes (P5 target T3 F1${\geq}0.40$ on $12$
zero-training classes), geospatial priors expected
to reduce routing path length in peak-risk
conditions, and geo-tagged references that drive
image-to-image overconfidence detection. \textbf{\observe}
targets ECE ${\leq}0.13$ on \pv ($54{,}306$
controlled images, zero training overlap) and
${\leq}0.18$ on the complete \pw benchmark, with the
hypothesis that the student surpasses its
\plantswarm teacher on both --- at a training-to-test
ratio of ${\approx}1{:}280$. The numbers in this
draft come from a smoke pilot on Tomato + Soybean
(Appendix~\ref{app:smoke}); the production run that
adjudicates the seven pre-registered predictions of
Appendix~\ref{app:predictions} is pending.

The question this system answers is not whether
AI can diagnose plant disease with enough data.
The question is what it can do with the data
that actually exists---$10$ field photographs,
GPS coordinates, and a mechanistic understanding
of how pathogens work.

\section*{Ethics Statement}

\bugwood images are publicly available under
academic and extension-service terms.
GPS coordinates are retained at agro-ecological
zone granularity in \pathomedb, not at
precision levels that could identify individual
farms or farmers. Treatment recommendations
are informational only; deployment requires
trained extension workers with local regulatory
knowledge. We acknowledge \bugwood's geographic
bias toward North America and Europe and release
Layer~3 regional coverage statistics alongside
\pathomedb so users can assess reliability
before deployment. Per-disease ECE is released
alongside accuracy metrics.

\section*{Limitations}

\bugwood geographic bias limits Layer~3 accuracy
for Sub-Saharan Africa and South Asia.
Layer~5 reference library is small ($78$ images);
rare disease stages may retrieve suboptimal references.
\observe calibration thresholds require recalibration
under substantial distribution shift.
$\kappa$ circularity partially propagates label noise.
Zero-shot evaluation covers $12$ disease classes;
broader coverage strengthens the claim.

\section*{Acknowledgments}
[Omitted for anonymous review.]

\bibliographystyle{plainnat}
\bibliography{pathome3}

%% ---------------------------------------------------------------
\appendix

\section{Bugwood CSV Filtering}
\label{app:selection}

The IPMNet export
(\texttt{BugWood\_Diseases.csv}, $19{,}749$ rows)
is normalised by
\texttt{scripts/filter\_bugwood\_csv.py} in
five steps:
\begin{enumerate}
  \item drop rows with empty \texttt{Host Name}
    ($2{,}171$);
  \item map \texttt{Host Name} via
    \texttt{BUGWOOD\_EXACT\_CROP\_MAP}---a
    $60{+}$-entry crop synonym table imported
    from the unified \texttt{DataLoader.py}
    registry; non-host taxa (``wood decay
    fungi'', ``shelf fungi'', \ldots) are
    dropped;
  \item strip the parenthetical Latin binomial
    from \texttt{Subject Display Name} (regex
    \texttt{\textbackslash{}s*\textbackslash{}(.*\$})
    and title-case the result;
  \item drop rows whose \texttt{Location}
    (state name) is empty or unrecognised;
  \item group by
    $(\text{NormCrop},\,\text{NormDisease})$
    and keep classes with $\geq{10}$ candidate
    rows.
\end{enumerate}
Output: \texttt{BugWood\_Diseases\_usable.csv},
$11{,}513$ rows over $\mathbf{484}$ classes,
plus five derived columns (\texttt{NormCrop},
\texttt{NormDisease}, \texttt{StateLat},
\texttt{StateLon}, \texttt{AezCode}).

The first $10$ rows per class (deterministic
ordering by \texttt{Image Number}) are admitted
by \texttt{BugwoodLoader}; $7$ feed \plantswarm
trace generation ($30$ runs each,
temperature $T{=}0.9$, max path length $15$
steps, backtrack budget $1$ per episode); $3$
are held as references for the
\texttt{ReferenceLibrary}.
\texttt{select\_diverse\_subset}
($k$-medoids over CLIP ViT-B/32 embeddings) is
still exposed for callers who want diversity
sampling within a class's pool, but the
deterministic Image-Number ordering is
sufficient for reproducibility.

\section{Geo Prior at State Granularity}
\label{app:layer3}

The geo prior is computed directly off
\texttt{state\_counts} on each
\texttt{SymptomProfile}: there is no separate
regional-epidemiology table to maintain.
US state names from the CSV's \texttt{Location}
column are resolved to a population-weighted
centroid
(\texttt{utils.geo.US\_STATE\_CENTROID}, $50$
states $+$ DC $+$ US territories that
occasionally appear in IPMNet) and forwarded
to the FAO AEZ lookup. State cells with
$<3$ observations are flagged \texttt{sparse}
and \observe queries
\texttt{db.symptoms.global\_prior(d)} in
those cases. The earlier formulation's
AEZ-month grid and EPPO Pearson-$r$ validation
were dropped because the IPMNet CSV provides
neither the GPS nor the capture-date fields
needed to construct them; the coarse-fallback
AEZ table maps the entire US footprint into
two zones (TMP, STM), and the full
$\sim$$17$-zone resolution requires
\texttt{PATHOME\_AEZ\_SHAPEFILE} pointing at
a real FAO GAEZ shapefile.

\section{Training Configuration}
\label{app:training}

\textbf{Phase A (DT):} AdamW $\text{lr}{=}10^{-4}$,
cosine decay, warmup $500$ steps, batch $32$
($8{\times}4$ accumulation), $50$ epochs,
early stopping patience $5$ (validation ECE).
Train/val/held-out split is grouped \emph{by source
image\_id} (stripping the \texttt{::run<NN>} suffix
before partitioning) so all $30$ runs of a given
\bugwood image stay in the same fold and intra-image
leakage is impossible. At production scale, the
80/10/10 partition is approximately
$145/18/19$ images $\times 30$ runs
$\approx 4{,}350/540/570$ traces. Previous drafts
reported a per-trace partition of $4{,}368/546/546$;
the per-image partition supersedes it as of the
code change recorded in Appendix~\ref{app:smoke}.
1$\times$ A100 40GB, ${\approx}4$--$6$h.

\textbf{Phase B (GRPO):} AdamW $\text{lr}{=}5{\times}10^{-5}$,
$G{=}8$ rollouts, $10$ epochs, $\beta_\text{KL}{=}0.04$,
clip $\epsilon{=}0.2$, reference policy frozen from Phase~A.
1$\times$ A100 40GB, ${\approx}3$--$4$h.

\section{Falsifiable Predictions}
\label{app:predictions}

\begin{table}[H]\centering\small
\begin{tabular}{lp{3.2cm}c}
\toprule
\textbf{P} & \textbf{Quantity} & \textbf{Target} \\
\midrule
P1 & $\rho(L, H[\hat{y}])$ & $+0.42$--$+0.55$ \\
P2 & $\Delta$Acc, PathogenAgent post-BT & $+9$ F1 \\
P3 & acc(early) $-$ acc(extended) & $+12$ F1 \\
P4 & \observe ECE (\pv / \pw) & ${\leq}0.13/0.18$ \\
P5 & Zero-shot T3 F1 (12 unseen classes) & ${\geq}0.40$ \\
P6 & Cross-crop T3 F1, iter.\ 3 & ${\geq}80\%$ \\
P7 & Layer~3 vs EPPO Pearson $r$ & ${\geq}0.70$ \\
\bottomrule
\end{tabular}
\caption{Seven falsifiable predictions. P5 and P7
  are new relative to prior formulations.}
\label{tab:predictions}
\end{table}

\section{Smoke Pilot: Tomato + Soybean (2026-05-11)}
\label{app:smoke}

The numbers reported in this draft are from a
\emph{smoke pilot} of the full pipeline rather than the
production run. The smoke pilot is intended to validate every
code path with minimal compute before we commit to the
full Phase~0 budget (${\approx}\$50{-}150$ in Anthropic API
quota) and the ${\approx}60$~h of A100 time for Phases~2 and~4.
The configuration restricts Bugwood to two crops
(\textit{Tomato} and \textit{Soybean}) and downscales every
budget knob; everything else --- the canonical/regional KB
schema (§\ref{sec:pathomedb}), the routing trace format
(§\ref{sec:plantswarm}), the OBSERVE training objective
(§\ref{sec:observe}), and the eval harness --- is run
unmodified.

\paragraph{Scope.} 2 crops $\times$ 25 (crop, disease) classes
after threshold ${\geq}15$ filter: 14 Tomato, 11 Soybean.

\paragraph{Image-disjoint train / val / test splits.}
\begin{table}[H]\centering\small
\begin{tabular}{l l r}
\toprule
\textbf{Split} & \textbf{Source} & \textbf{Images} \\
\midrule
Train (Bugwood)    & \texttt{split="trace"} & 75 \\
Val   (Bugwood)    & \texttt{split="val"}   & 25 \\
Test  (PlantVillage) & \texttt{split="test"} & 200 \\
Test  (PlantWild)  & \texttt{split="test"}   & 200 \\
\bottomrule
\end{tabular}
\caption{Smoke-pilot splits. Train and val are sliced
deterministically by \texttt{Image Number} ascending after
per-class dedup; the loader's slice ($\texttt{[:3]}$ vs
$\texttt{[3:4]}$) yields image-disjoint sets by construction,
audited at load via
$\mathrm{set}(\textsc{train}) \cap \mathrm{set}(\textsc{val}) = \emptyset$.
Test comes from separate datasets whose image-id schemes
cannot collide with Bugwood's \texttt{bugwood::N}.}
\label{tab:smoke-splits}
\end{table}

The previous formulation re-used the second slice as the
PathomeDB Layer-5 CLIP exemplar pool, which leaked held-out
images into Phase~2 trace-time retrieval. The smoke pipeline
disables Layer-5 retrieval entirely
(\texttt{reference\_records=[]}) so the val split never
shows up as a retrieval hit at any stage.

\paragraph{Downscaled budgets vs production.}
\begin{table}[H]\centering\small
\begin{tabular}{l r r}
\toprule
\textbf{Knob} & \textbf{Production} & \textbf{Smoke} \\
\midrule
Classes & 484 & 25 \\
\texttt{per\_class} / \texttt{trace\_split} & 10 / 7 & 4 / 3 \\
\texttt{runs\_per\_image} & 30 & 3 \\
Total traces & 5{,}460 & 225 \\
\texttt{Tmax} & 15 & 8 \\
DT epochs & 50 & 3 \\
GRPO epochs & 10 & 1 \\
LoRA rank & 16 & 8 \\
\pv\ / \pw\ eval & 54{,}306 / 18{,}000 & 200 / 200 \\
\texttt{bootstrap\_n} & 1{,}000 & 100 \\
\bottomrule
\end{tabular}
\caption{Smoke knobs (\texttt{smoke/bugwood\_pathome\_smoke.yaml}).}
\label{tab:smoke-budgets}
\end{table}

\paragraph{Phase~4 split is per source-image.} The 80/10/10
partition of trace records is grouped by source
\texttt{image\_id} (stripping the \texttt{::run<NN>} suffix)
before splitting, so all three runs of a single Bugwood image
stay in the same fold. The prior per-trace split allowed three
runs of one image to land in train, val, and held
simultaneously --- a silent intra-image leakage path. The new
partition asserts non-overlap on the unique image set at load
time.

\paragraph{Runtime and cost.} Phase~0 (Claude-headless,
canonical KB + per-state visual deltas) runs locally,
${\approx}45{-}90$~min for ${\approx}\$5{-}15$ in API quota.
Phases~1--5 (build, trace, enhance, train $\times 2$,
eval $\times 4$) run as a single A100 job on Nova,
${\approx}60{-}90$~min total walltime. The seed file
(\texttt{smoke/artifacts/pathome\_seed/symptoms\_seed.json},
25 \texttt{SymptomProfile}s with canonical summaries,
treatments, and per-state deltas) is the only Phase~0 output
that has to cross the LOCAL${\to}$NOVA boundary; everything
else is reproducible on Nova from the committed code.

\paragraph{What this pilot validates.} The smoke run
exercises every code path that the production run will hit
--- the SAGE port to \texttt{pathome\_kb}, the
canonical+regional schema round-trip, the trace JSONL writer
with its per-image atomic JSON mirror, the per-image OBSERVE
split, and the eval harness --- on $\sim 1/19$ of the
images and $\sim 1/24$ of the trace volume. The numbers
reported here are not statistically meaningful (eval $n=200$,
1~GRPO epoch); the contribution of this pilot is to show that
the pipeline producing them ran end-to-end without code-path
regressions before committing to the full spend.


% ============================================================
% Auto-generated artefacts. Filled in by:
%   scripts/viz/kb_stats.py        -> auto_kb_stats.tex
%   scripts/viz/observe_curves.py  -> auto_observe_curves.tex
%   scripts/viz/observe_eval.py    -> auto_observe_eval.tex
%   scripts/viz/trace_stats.py     -> auto_trace_stats.tex
% The viz scripts also emit PNGs under results/figures/, which the
% LaTeX \graphicspath above resolves automatically.
% Run `bash scripts/viz_all.sh` after the pipeline finishes to refresh.
% ============================================================

\section{Auto-generated Implementation Results}
\label{app:auto-results}

The figures and tables below are regenerated from each
end-to-end run by \texttt{bash scripts/viz\_all.sh}; the
LaTeX snippets land at \texttt{plantswarm/latex/auto\_*.tex}
and the PNGs at \texttt{results/figures/}.

\subsection{PathomeDB seed summary}
\IfFileExists{auto_kb_stats.tex}{\input{auto_kb_stats}}{%
  \emph{(KB stats not yet generated; run
  \texttt{bash scripts/viz\_kb.sh}.)}}

\subsection{Phase 0R routed-swarm trace stats}
\IfFileExists{auto_trace_stats.tex}{\input{auto_trace_stats}}{%
  \emph{(Trace stats not yet generated; run
  \texttt{bash scripts/viz\_traces.sh}.)}}

\subsection{\observe training curves}
\IfFileExists{auto_observe_curves.tex}{\input{auto_observe_curves}}{%
  \emph{(\observe training history not yet generated; run
  \texttt{bash scripts/viz\_observe.sh}.)}}

\subsection{\observe held-out evaluation}
\IfFileExists{auto_observe_eval.tex}{\input{auto_observe_eval}}{%
  \emph{(\observe eval not yet generated; run
  \texttt{bash scripts/viz\_observe.sh}.)}}

\end{document}
